---
title: "HGA data for Analysis of Hops Yield"
author: "Don A. Lloyd\n"
date: "Updated `r format(Sys.time(), '%d %B, %Y')`"
output: github_document
# always_allow_html: true
editor_options: 
  chunk_output_type: console
---

```{r knit, message=TRUE, warning=TRUE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<a name="top"></a> Keywords: hops, yields, data extraction, data cleansing, data validation, regex, outlier analysis

[Extracting](#extracting-hop-yields-from-hga) and [cleaning](#data-cleaning) hop yields from HGA reports

[Validating the yield data with NASS](#validating-the-hga-yield-data-with-nass) including [internal checks](#validate-nass-internal-reporting)

<!-- [GO TO WORK IN PROGRESS](#in-progress-data-checks) -->

[Finalizing the yield dataset](#finalizing-the-yield-dataset)

<!-- outlier tests, continuity checks -->
[Preliminary analysis of hops yields](#preliminary-analysis-of-hop-yield-series) including [outliers](#outliers) and series [continuity](#continuity)

Write the [final yields](#final-yields) 
<!-- and demonstrate basic [stationarity tests](#stationarity-tests) of the series -->

[Summary](#summary) and some [challenges](#challenges-with-yield-data)

[Session info](#session-info-and-notes)

```{r libs, echo=TRUE, warning=FALSE, message=FALSE}
library(here)
library(tictoc)
library(dplyr)
library(stringr)
library(tidyr)
library(tibble)
library(pdftools)
library(ggplot2)
library(tseries)
library(data.table) # fwrite
library(lubridate) # today
library(keyring)
library(rnassqs) # NASS API
```

```{r setup}
HGA_VERBOSE = FALSE
HGA_DATA_PATH = "~/Dropbox/Data/HGA"
SUBDIR = "yield"
```

## Extracting hop yields from HGA

In 2020, wildfires caused extensive smoke damage to crop harvests in and around Oregon's Willamette River Valley. The smoke damage to wine grapes were widely reported. But, I enjoy a good IPA more often than a good pinot so I became interested in how hops harvests were impacted by the wildfires. The number and severity of wildfires in the Willamette Valley that year was worsened by high temperatures and dry conditions. I began looking for hops cultivation data to understand several questions:

* Do we see evidence that hop yields negatively impacted in 2020?

* What environmental observations -- if any -- correlate with yields or changes in yields?

* Are some hops varieties or growing regions more or less susceptible to changes in environmental variables?

The overall strategy for conducting this analysis has been to compile climate and air quality data for areas in the PNW that grow hops. Sourcing and processing of environmental data, and finding hop cultivating areas are described in other scripts.

To begin the analysis, I compiled hops yield data from statistical reports published annually by the [Hop Growers of America](https://usahops.org) since 2000. I wrote much of the scripting in this document to process the reports by identifying and parsing the yield tables from each pdf and correct transcription issues. 

I then used the [National Agricultural Statistics Service](https://nass.usda.gov) (NASS) to validate my transcription of the data and update missing values. NASS is an excellent source of agricultural data in the US but it lacks hops reporting by variety before 2013. Validating against NASS was a very useful exercise for improving data cleaning in my pdf processing. 

The yield tables include annual yields by variety and state, and summary values by state, region, and or as national measures. None of the available yield data provide subtotals by county in which hops were grown.

I am using the HGA yield values when first published, not most recently published. Each HGA table also reports several prior year's values to indicate changes in reporting and recent trends. Values from later reports for a given year could be used to test for changes in reporting, for example to revise a previously published value.

This script produces three csv files reporting hops yields in three levels of granularity: within the PNW region, by state within the PNW, and by variety within each state. Each file reports Year, Yield, and data Source. State and variety labels are included in the respective files. Outlying values in each data set are identified with dummy variables.

The script also produces a csv of validation data from NASS, and a table of reporting continuity for varieties grown in each state.

You can download the data files from this git and get on with your own analysis. If you choose to download the PDFs from HGA just save them in a directory of their own and define the full `HGA_DATA_PATH`. 

### PDF processing

```{r HGA_process_PDFs}
# this relies on lots of regex to find and parse yield tables that
# appear on different pages of each annual report. there are lots of
# issues to work around in these reports including changing headers,
# misplaced labels, tables split across multiple pages, footnote markers
# embedded in labels, variable spacing of raw table data, and 
# inconsistent labeling in different reports...
# use HGA_VERBOSE=TRUE for debugging 
#
pdf.list <- dir(HGA_DATA_PATH, pattern = ".pdf", ignore.case = TRUE, full.names = TRUE)

tic() # for the curious
# read each report, find the yield table and extract it
listtbl <- lapply(pdf.list, function(x) {
  if (HGA_VERBOSE) cat(sprintf("processing %s\n", basename(x)))
  rawtext <- pdf_text(x) %>% str_split("\n")
  singlepage <- lapply(rawtext, function(x) Reduce(c, x))
  # find header lines
  hp <- grep("[[:digit:]]{4} Statistical .*", singlepage)
  hl <- grep("[[:digit:]]{4} Statistical .*", singlepage[[min(hp)]])
  # extract the reporting year
  year <- as.numeric(gsub(".*[ ]+([[:digit:]]{4}) Statistical .*", "\\1", 
                          grep("[[:digit:]]{4} Statistical .*", singlepage[[min(hp)]], value = TRUE)
  ))
  out <- NULL
  # recent reports refer to "PNW HOP YIELDS", older reports use "U.S. HOP YIELDS"...
  ii <- grep("HOP YIELDS BY STATE", singlepage, ignore.case = TRUE)
  # find page in report that includes the current yield table by variety
  toc <- grep("HOP YIELDS.*VARIETY", unlist(singlepage[ii[1]]), ignore.case = TRUE, value = TRUE)
  pptxt <- gsub(".* ([0-9\\-]+)$", "\\1", toc)
  if (grepl("\\-", pptxt)) pptxt <- unlist(strsplit(pptxt, "\\-"))
  # can test if pp is defined
  pp <- as.numeric(pptxt)
  # isolate page numbers
  pagenumtarget = "[[:space:]]{5,}([[:digit:]ivx]+).*"
  pagenum <- sapply(1:length(singlepage), function(x) {
    pagerows <- grep(pagenumtarget, singlepage[[x]])
    if (length(pagerows) >0) {
      thispage <- max(pagerows)
      gsub(pagenumtarget, "\\1", singlepage[[x]][thispage])
    }
  })
  pp <- as.numeric(match(pptxt, pagenum))
  o <- trimws(unlist(singlepage[pp]))
  # A few files will add breaks before and or after TM labels in Variety names
  # sometimes leaving a row of numbers with no leading Variety name. We look
  # for the missing data in lines immediately preceding or following each row
  # missing a hop variety.
  # let's first eliminate "Hop Grower" lines
  o <- grep("^Hop Growers", o, invert = TRUE, value = TRUE)
  # then drop rows that are at least 5 char all caps (typically state names and some headings)
  o <- grep("^[[:upper:]]{5,}$", o, invert = TRUE, value = TRUE)
  # look for rows that start with a numeric (e.g. yield) entry that are otherwise complete
  # these rows are missing a variety entry
  # this is a little cumbersome, but better than other methods I have tried...
  ifloat <- which(grepl("^[0-9,]{3,}|^TM[ ]+|^[12]{1}[ ]*", o) & nchar(o) >30)
  if (length(ifloat) >0) {
    if (HGA_VERBOSE) {
      cat("these rows appear to be missing variety labels\n")
      print(ifloat)
    }
    # check previous 2 rows for varietal label that got left behind
    for (i in ifloat) {
      if (HGA_VERBOSE) cat(sprintf("checking line %d:\n%s\n", i, o[i]))
      ivar <- c(i-2, i-1)
      # in rare cases the pdf publishing software will format the
      # variety label by shifting it up or down one row, so let's fix it
      # the logic is a little pedantic, but easy enough to read
      if (match(i, ifloat) < length(ifloat)) {
        ok <- (i+1 < ifloat[match(i, ifloat) +1] -2)
      } else {
        ok <- (i+1 < length(o))
      }
      if (ok) ivar <- c(ivar, i+1)
      if (HGA_VERBOSE) print(o[sort(c(ivar,i))])
      ii <- which(nchar(o[ivar]) <16)
      if (length(ii) >0) {
        # replace corrected line using wide space so it parses with remainder of table
        # n.b. the partial lines we built labels from won't be parsed below...
        if (HGA_VERBOSE) cat(sprintf("proposed label: %s\n", paste(o[ivar[ii]], collapse = " ")))
        o[i] <- paste(paste(o[ivar[ii]], collapse = " "), o[i], sep = "          ")
        if (HGA_VERBOSE) print(o[i])
      } else {if (HGA_VERBOSE) cat("no label text found...\n")}
    } 
  } # closes loop for missing label search
  # find rows that conform to a table definition
  rawtbl <- grep("([0-9,\\.]+|[\\*]){3,}", o, value = TRUE) %>%
    # replace blank space with a delimiter unique to the table
    gsub("[[:space:]]{2,}", "|", .) %>% 
    gsub(",|%", "", .) %>% # drop comma and percentage formatting
    # next line partially fixes issues with extra spaces or footnote labels in first col
    gsub("([[:alpha:] \\!]+)\\|([[:alpha:] ]+|[123]?\\|.*)", "\\1 \\2", .)
  tblrows <- grep("\\|", rawtbl) # keep only lines we can parse
  rawtbl <- rawtbl[tblrows]
  # find header row
  h <- grep("Variety", rawtbl, ignore.case = TRUE)
  # header may be duplicated if table continues on second page
  # find start of next table
  h1 <- min(h)
  # n.b. this header appears to be consistent across the reports
  hdr <- unlist(strsplit(rawtbl[h1], split = "\\|"))
  # if there are three or more header rows with a split table, assume
  # they mark new tables that can be dropped...
  hh <- ifelse(length(h) >2, h[3] - 1, length(rawtbl))
  rawtbl <- rawtbl[1:hh] # dropping extra rows
  # eliminate stray headers
  rawtbl <- grep("& Variety", rawtbl, ignore.case = TRUE, value = TRUE, invert = TRUE)
  # ready to parse what is left
  # read.table orig. had skip = h1, error caused us to skip several years of
  # single line Idaho reporting...
  parsedtbl <- read.table(text = rawtbl, skip = 1, sep = "|", fill = TRUE,
                          na.strings = c("*"), row.names = NULL,
                          col.names = hdr) %>%
    as_tibble %>%
    select(1, all_of(paste0("X", year))) %>%
    rename(Variety = 1, Yield = 2) %>%
    mutate(Yield = ifelse(grepl("[^[:digit:]]", Yield), NA, Yield)
           ,Yield = as.numeric(Yield)
           ,Year = year
           ,isTotal = grepl("Total", Variety)
           ,Variety = gsub("Total Idaho[ ]*1", "Total Idaho", Variety)
           ,State = ifelse(isTotal, trimws(gsub("Total", "", Variety)), NA)
           ,Source = paste("HGA", basename(x))
    ) %>%
    fill(State, .direction = "up")
  list(file = basename(x), tbl = parsedtbl, year = year, 
       raw = rawtbl, o = o)
}) # next file
toc()

hga_years <- unlist(lapply(listtbl, function(x) x$year))
listtbl <- setNames(listtbl, hga_years)

# save the years of yield reporting to correctly bound
# climate and air data sets later on
# fwrite(data.frame(years = sort(years)), file = here("HGA_years.csv"))
```

The main disadvantage of compiling the first reported yield for each variety or aggregate is that we can overlook subsequent revisions to the reporting. Inspection of consecutive reports shows that yields can be revised later, usually in the next year. Many yields that are later corrected may be preliminary estimates.

[Back to top](#top)

### Data cleaning

Cleanup on names seems self evident to me, but for background and explanatory info see: https://en.wikipedia.org/wiki/List_of_hop_varieties

For naming conventions see https://www.hopslist.com/hops/

Abbreviations for some of the major hop breeders:\n

* YCR = Yakima Chief Ranches, consortium of family hop farms, Yakima, WA
* HBC = Hop Breeding Company, a joint venture of YCR and Haas 
* VGX = Virgil Gamache Farms, Toppenish, WA
* ADHA = Association for the Development of Hop Agronomy (www.adha.us)

_Some variety names are trademarked._ I have attempted to capture the trademark status while parsing the HGA reports.

```{r HGA_Yield_Table}
# process each table, clean and parse variety names
# I do attempt to preserve TM and R notations...
hga_yield <- 
  lapply(listtbl, function(x) x$tbl) %>% 
  bind_rows() %>%
  filter(!str_detect(Variety, "TATE")) %>%
  # the 2012 report does not include the total rows for WA or US/PNW
  mutate(State = ifelse(is.na(State) & Year == 2012, "Washington", State)
         ,isTM = grepl("TM|T M|\\x99", Variety) # trademark
         ,isR = grepl(" R|R |\\xAE", Variety) # registered trademark
         # eliminate commas, parentheses and symbols from variety names
         ,Variety = gsub(",|\\(|\\)|TM|T M|\\x99|\\xAE|\\*", "", Variety)
         ,Variety = trimws(Variety) # clean up trailing spaces after symbol pass
         # ,Variety = gsub(",|\\(|\\)|TM|T M| R|R |\\x99|\\xAE", "", Variety)
         # contract multiple spaces
         ,Variety = gsub("[[:space:]]{2,}", " ", Variety)
         # eliminate strays spaces within variety names
         ,Variety = gsub("([[:alpha:]]+)[[:space:]]+([[:lower:]]+.*)", "\\1\\2", Variety)
         # drop subscripts from total and aggregate lines
         ,Variety = gsub("(Total|Other [A-Za-z]+)[ ]*[12]$", "\\1", Variety)
         # drop -er from Hallertauer, see NASS labels
         # ,Variety = gsub("Hallertauer", "Hallertau", Variety)
         ,Variety = gsub("Calyp so", "Calypso", Variety) # unusual space
         # standardize variety names to correct some inconsistencies 
         # but also make them more consistent with NASS during validation
         ,Variety = case_when(
           grepl("YCR[ ]*4|Palisade", Variety) ~ "Palisade YCR 4"
           ,grepl("YCR[ ]*5|Warrior", Variety) ~ "Warrior YCR 5"
           ,grepl("Super[ ]*Galena", Variety) ~ "Super Galena"
           ,grepl("Talus|692", Variety) ~ "Talus HBC 692"
           ,grepl("Summit", Variety) ~ "Summit"
           ,grepl("Ahtanum|YCR 1$", Variety) ~ "Ahtanum YCR 1"
           ,grepl("ADHA-483|Azacca", Variety) ~ "Azacca ADHA-483"
           ,grepl("ADHA-881|Jarrylo", Variety) ~ "Jarrylo ADHA-881"
           ,grepl("Pekko|ADHA-871", Variety) ~ "Pekko ADHA-871"
           ,grepl("Citra|HBC 394", Variety) ~ "Citra HBC 394"
           ,grepl("Tettnang", Variety) ~ "Tettnanger"
           ,grepl("Loral|HBC 291", Variety) ~ "Loral HBC 291"
           ,grepl("Simcoe|YCR 14", Variety) ~ "Simcoe YCR 14"
           # Hallertau Magnum was reported as Magnum from 2012-2018
           # but review of HGA yield tables show their equivalence
           # n.b. Hallertau Magnum and Hallertauer are reported as distinct varieties
           #,grepl("Magnum", Variety) ~ "Hallertau Magnum"
           ,TRUE ~ Variety
         ) 
  )

fwrite(hga_yield, here(SUBDIR, "hga_yield.csv"))

# hga_yield table retains "Other" varieties and totals
(hga_varieties <- 
    hga_yield %>%
    pull(Variety) %>%
    sort %>%
    unique
)
```

Get our bearings for what typical yield values (lb/acre) we should expect. I am leaving in the aggregations for state and region with individual varieties to get a ballpark figure over the entire reporting period. Yields of 1500-2000 lb/acre are typical. 

We should also note that we have over 200 missing yields transcribed from the HGA reports. These are anonymized values to protect the privacy of growers that are cultivating unique varieties within their state.

```{r hga_yield_basics}
summary(hga_yield$Yield)
ggplot(hga_yield, aes(Yield)) + geom_histogram(bins=20, fill="forestgreen")
```

UPDATE

The missing national and WA aggregate yields from 2012 are truly missing from the HGA data, not merely skipped in our processing. Add the records, label their source, and illustrate the results. We then subset and save the yield tables into state and regional summaries, and varietal yields by state.

[Back to top](#top)

## Validating the HGA yield data with NASS

To validate, we can retrieve hops yields from the National Agriculture Statistics Server (NASS). This pulls all hops data that will overlap the our HGA reporting years, up to the current year. Since hops are harvested once per year, statistics reported for the current year are usually incomplete before December.

The data returned includes yields (lbs/acre), harvested acres, and production in pounds. The data also includes production measured in dollars and many categorical factors that are not immediately relevant to this analysis. 

```{r HGA_Validation_1_NASS_Table}
key_get("NASS") %>%
  nassqs_auth()

# n.b. can't yet select on prodn_practice_desc
nass_all <- nassqs(
  source_desc = "SURVEY"
  ,sector_desc = "CROPS"
  ,commodity_desc = "HOPS"
  ,year__GE = min(hga_years) # 20250826 updated API
  # adding a second constraint on year doesn't seem to work
  #,year__LT = year(today()) 
  ,reference_period_desc = "YEAR"
  ,progress_bar = FALSE
) %>%
  # exclude current year, assume its reporting is incomplete
  # yields and other data are not usually finalized until December
  filter(as.numeric(year) < year(today()))

fwrite(nass_all, file = here(SUBDIR, "nass_all_data.csv"))
```

### Validate NASS internal reporting

We can start by testing whether the NASS yields are internally consistent with the corresponding production and acreage values. After isolating production and area in the correct units, we can calculate the yields from their ratio and look for discrepancies with the reported yields.

```{r nass_internal_validation}
nass_internal_validation <-
  nass_all %>%
  # prodn_practice_desc now tracks ORGANIC as a subset of production
  filter(prodn_practice_desc != "ORGANIC") %>% # we want totals regardless of production practice
  filter((statisticcat_desc == "YIELD" & unit_desc == "LB / ACRE") | 
           (statisticcat_desc == "PRODUCTION" & unit_desc == "LB") |
           (statisticcat_desc == "AREA HARVESTED" & unit_desc == "ACRES")
  ) %>%
  pivot_wider(id_cols = c("state_name", "class_desc", "year"), 
              names_from = "statisticcat_desc", values_from = "Value") %>%
  filter(!is.na(YIELD)) %>%
  mutate(`YLD CALC` = round(PRODUCTION /`AREA HARVESTED`, 0)
         ,ERR = `YLD CALC` -YIELD
         ,PERC_ERR = ERR /YIELD *100)

# we will ignore rounding errors of +/- 1 lb/acre
filter(nass_internal_validation, abs(ERR) >1) %>%
  rename(ACRES = `AREA HARVESTED`) %>%
  arrange(desc(abs(ERR)))
```

There are two very minor errors (< 1%) that we will ignore by preferring the reported yield over the calculation. The other errors are quite large and difficult to diagnose for the "OTHER VARIETIES" label. The errors are most likely a consequence of the aggregation or incomplete reporting of harvested acres. Whatever the source, the calculated yields are obvious departures from the ranges we might expect from our summary of HGA data, and we will use the reported yields for each discrepant case. 

```{r nass_yield}
# annual yields by state per NASS
# class_desc distinguishes yields are aggregated over variety by region
# class_desc == "ALL CLASSES" are state/PNW/US totals
# class_desc != "ALL CLASSES" are varietal yields

nass_yield <- 
  nass_all %>%
  tibble %>%
  mutate(Year = as.numeric(year)) %>%
  filter(statisticcat_desc == "YIELD"
         #,class_desc != "ALL CLASSES"
         ,prodn_practice_desc != "ORGANIC"
  ) %>%
  select(class_desc, Value, Year, state_name) %>%
  mutate(Yield = ifelse(grepl("\\(D\\)", Value), NA, gsub(",", "", Value))
         ,Yield = as.numeric(Yield)
         ,Variety = gsub(",| TM| R[^[:upper:]]| R$", "", class_desc)
         ,Variety = case_when(
           Variety == "MILLENIUM" ~ "MILLENNIUM" # fix persistent NASS typo
           ,grepl("HBC 682", Variety) ~ "PAHTO HBC 682" # NASS doesn't use Pahto name
           # ,grepl("C/T/Z", Variety) ~ "COLUMBUS/TOMAHAWK" # HGA still use C/T label (16 June 24) -- will reconcile this below
           ,grepl("SABROHBC 438", Variety) ~ "SABRO HBC 438" # typo (16 June 24)
           # next line formerly known as X-331 from an OR exp. hop field
           ,grepl("STRATAOR91331", Variety) ~ "STRATA OR 91331" # (5 Sep 25 expand collapsed text)
           ,TRUE ~ Variety
         )
         ,State = str_to_title(state_name)
  ) %>%
  select(Variety, Yield, Year, State) %>% 
  arrange(Year, State)

(nass_varieties <- 
    nass_yield %>%
    pull(Variety) %>%
    sort %>%
    unique
)

fwrite(nass_yield, file = here(SUBDIR, "nass_yield.csv"))
```

[Back to top](#top)

## Finalizing the yield dataset

We have already standardized variety names _within_ each data set, but before combining data we should examine mismatched labels across the two data sets. As an aside, I found that investigating discrepancies between HGA and NASS was very instructive for troubleshooting pdf processing errors, and surfaced some curious issues with the 2013 reporting.

```{r precombine_label_checks}
var_hga <- pull(hga_yield, Variety) %>% unique %>% toupper # %>% sort
var_nass <- pull(nass_yield, Variety) %>% unique # %>% sort

# varieties in NASS not found in HGA
setdiff(var_nass, var_hga) %>% sort

# varieties in HGA not found in NASS
setdiff(var_hga, var_nass) %>% sort
```

Start by examining the Hallertauer varieties. Let's try matching yields across the two data sets to look for equivalence among different labels.

```{r hallertau_tests}
inner_join(hga_yield %>%
             select(-starts_with("is"), -Source) %>%
             filter(grepl("Hallertau|Magnum", Variety), !is.na(Yield))
           ,nass_yield %>%
             filter(grepl("HALLERTAU", Variety), !is.na(Yield))
           ,by = c("Year", "State", "Yield"), suffix = c("_HGA", "_NASS")
           )
```

Now we can address three labeling issues before combining the yields.

* NASS assigns state and region aggregate values to "ALL CLASSES", which I will re-label to match HGA aggregates.

* HGA spell out Columbus/Tomahawk/Zeus while NASS uses the abbreviation C/T/Z. I choose to spell out the names to make it little easier to compare with other Columbus/Tomahawk and Zeus varieties.

* NASS use the label HALLERTAUER MITTELFRUHER where HGA just uses Hallertauer. And while we already cleaned up the HGA labels for Magnum, we see that NASS uses HALLERTAUER instead of HALLERTAU for that variety. We can most clearly see the equivalence through matching the yields for each. I am going to defer to the NASS choices for both.

I'll apply these corrections to the NASS data during the merge. Then,
having standardized variety names across the data sets, we can examine the combined yield data sets. We have to expect other and experimental descriptors to be inconsistent over time as varieties in those categories change. We will keep state and region totals for separate analysis. Among groupings of variety, state, and year we can:

1. retain HGA where both NASS and HGA yields agree (e.g. < 1%)
2. analyze the discrepancy where both yields are reported but disagree
3. retain the single value where only one source reports the yield
4. report tuples that don't report a yield in either data set

```{r hop_yields}
tmp_yields <- 
  bind_rows(
    nass_yield %>%
      mutate(Source = "NASS"
             ,State = ifelse(grepl("Us", State), "United States", State)
             ,Variety = ifelse(Variety == "ALL CLASSES",
                               paste("TOTAL", toupper(State)),
                               Variety)
      )
      ,hga_yield %>%
      mutate(#Variety = case_when(
        # grepl("Magnum", Variety) ~ "Hallertau Magnum"
        # ,TRUE ~ Variety),
        Variety = toupper(Variety)
        ,Source = "HGA"
      ) %>%
      select(-starts_with("is"))
  ) %>%
  # ALPHA and AROMA are not varieties but generic attributes like OTHER
  # remove all generically labeled varieties
  filter(!grepl("ALPHA|AROMA|OTHER|EXPERIMENTAL", Variety)) %>%
  # NASS prefers the abbreviation, HGA spells out each name
  # there is no overlap between labels so we can standardize on either
  # I spell out the names to make it easier to compare with other
  # Columbus/Tomahawk and Zeus.
  mutate(Variety = ifelse(grepl("C/T/Z", Variety), "COLUMBUS/TOMAHAWK/ZEUS", Variety)
         ,Variety = gsub("HALLERTAU MAGNUM", "HALLERTAUER MAGNUM", Variety)
         ,Variety = gsub("HALLERTAUER$", "HALLERTAUER MITTELFRUHER", Variety)
         ) %>%
  # replace PNW total references with US
  mutate(Variety = gsub("PACIFIC NORTHWEST", "UNITED STATES", Variety)
         ,State = gsub("Pacific Northwest", "United States", State))
  # we haven't bothered to drop NA yields yet because pivoting to find
  # discrepancies will force NA for yields missing from either data set
  # use an anti-join to remove them from the data set

# look for discrepancies in yield reports
(yield_errors <- pivot_wider(tmp_yields, id_cols = c(Variety, State, Year),
                  names_from = Source, values_from = Yield) %>%
  mutate(ERR = HGA -NASS 
         ,PERC_ERR = round(ERR /HGA *100, 3)) %>%
  # filter(abs(ERR) >1) %>%
  filter(!is.na(PERC_ERR)) %>% #, abs(PERC_ERR) >0) %>%
  # keep NASS for 
  mutate(Drop = ifelse(abs(PERC_ERR) >0, "HGA", "NASS")) %>%
    arrange(desc(abs(PERC_ERR)))
)

fwrite(yield_errors, here(SUBDIR, "yield_errors.csv"))

yield_errors %>%
  mutate(abs_perc = abs(PERC_ERR),
         mag_err = case_when(
           abs_perc >1 ~ ">1%",
           abs_perc >0 ~ "<1%",
           TRUE ~ "IDENT")
  ) %>%
  pull(mag_err) %>%
  table

yield_errors %>%
  filter(abs(PERC_ERR) >0) %>%
  pivot_longer(cols = c(PERC_ERR, Year)) %>%
  ggplot(aes(x=value, group=name)) +
  geom_histogram(binwidth = 1) +
  facet_wrap(~ name, scales="free_x")
```

Most discrepancies are apparantly caused by revisions to HGA reporting. In those cases, the revised yields are already captured by NASS, which can be confirmed by comparing NASS to subsequent HGA reports.

The most common source of larger discrepancies are found in 2013, the first NASS reporting year. Its possible that NASS reporting in 2013 has transcription errors from a new import process that were corrected in later reporting. This may be worth investigating further, but for now I will retain the NASS values.

```{r knockout}
# create the knockout list
ko <- select(yield_errors, -NASS, -HGA, -ERR, Source=Drop)

hop_yields <- anti_join(tmp_yields, ko) %>%
  arrange(Variety, State, Year) %>%
  # finally drop missing yields
  filter(!is.na(Yield)) %>%
  # reassert the full HGA source filename for traceability
  left_join(.,
            hga_yield %>%
              filter(!is.na(Yield)) %>%
              select(, -starts_with("is"), -Yield) %>%
              mutate(Variety = toupper(Variety)
                     ,Detail = Source
                     ,Source = "HGA"
                     ,Variety = gsub("PACIFIC NORTHWEST", "UNITED STATES", Variety)
              )
  ) %>%
  mutate(Source = ifelse(!is.na(Detail), Detail, Source)) %>%
  select(-Detail) %>%
  arrange(State, Variety, Year)
# defer saving hops_yields until after outlier tests and 
# splitting by variety categories
```

[Back to top](#top)

## Preliminary analysis of hop yield series

### Outliers

Let's test for outliers using a very simple interquartile measure.

```{r outlier_test}
hops_outlier_test <-
  left_join(hop_yields,
            hop_yields %>%
              group_by(State, Variety) %>%
              summarize(q1 = quantile(Yield, p = 0.25)
                        ,q3 = quantile(Yield, p = 0.75)
                        ,iqr = q3-q1
                        ,t1 = q1 - 1.5*iqr
                        ,t3 = q3 + 1.5*iqr
              )
  ) %>% 
  # n.b. Group assignments will be used to name data sets later
  mutate(Group = ifelse(grepl("TOTAL", Variety), ifelse(State == "United States", "US", "State"), "Var")
         ,outlier = (Yield < t1) | (Yield > t3)
  )

n_out <- nrow(filter(hops_outlier_test, outlier))
cat(sprintf("%d outliers from IQR (%.1f%%)\n", n_out, n_out/nrow(hops_outlier_test) *100))

fwrite(hops_outlier_test, file = here(SUBDIR, "hops_outlier_test.csv"))

tot_yields <- select(hops_outlier_test, -q1, -q3, -iqr, -t1, -t3) %>%
  arrange(Variety, State, Year) 

tot_yields %>%
  filter(grepl("TOTAL", Variety), State != "United States") %>%
  ggplot(aes(x=Year, y=Yield, color=State)) +
  geom_line(linewidth = 0.8) +
  geom_point(data = filter(tot_yields, grepl("TOTAL", Variety), State != "United States", outlier),
             aes(x=Year, y=Yield), size = 2.2, show.legend = F) +
  geom_line(data = filter(tot_yields, State == "United States"),
            aes(x=Year, y=Yield), col = "black", linetype = "dashed") +
  theme(legend.position="bottom")
```

The full regional yield series is quite short, and most varietal series are even shorter. The two outliers shown above will likely be difficult to model regardless of method.

The PNW total yield (dashed line) follows the inflections of the Washington yields most closely of the three states. This just means the regional yield is weighted primarily by the larger Washington acreage committed to hops.

### Continuity tests

```{r continuity_test}
# this func. tests whether there are gaps in inclusively reported years
# the test for completeness relative to entire suite of data is performed separately
yield_runs <- function(yrs) {
  # yrs the years with observed yields for a given state/variety subset
  inclusive <- seq(min(yrs), max(yrs), 1)
  obs <- length(yrs)
  longest_run <- length(inclusive)
  longest_missing <- 0
  recent_end <- year(today()) -max(yrs) <2
  recent <- obs *recent_end
  if (length(inclusive) -length(obs) >0) {
    # convert the run of observation years into run of 0 (=not obs) and 1 (=obs)
    z <- rep(0, length(inclusive))
    z[match(yrs, inclusive)] <- 1
    run <- rle(z)
    ii <- which(run$values > 0)
    longest_run <- max(run$lengths[ii])
    jj <- which(run$values < 1)
    if (length(jj) >0) longest_missing <- max(run$lengths[jj])
    last_run <- length(run$values)
    # using the fact we assigned 1 to obs to implement two indicator funcs
    # so recent is an integer only if both the most recent years were observed 
    # and the most recent obs are within the past 2 years
    recent <- run$values[last_run] *run$lengths[last_run] *recent_end
  }
  tibble(start = min(yrs)
         ,end = max(yrs)
         ,obs = obs
         ,yrs = end -start +1
         ,longest_run
         ,longest_missing
         ,recent
  )
}

# test for completeness of series outside the yield_runs function
n_yr <- with(tot_yields, max(Year) -min(Year) +1)

(hops_continuity_test <- 
  tot_yields %>%
  reframe(yield_runs(Year), .by=c("Variety", "State")) %>%
  mutate(complete = (obs == n_yr))
)

fwrite(hops_continuity_test, file = here(SUBDIR, "hops_continuity_test.csv"))

# select yields by year for complete obs, ignoring aggregates
hops_continuity_test %>%
  filter(complete, !grepl("TOTAL", Variety)) %>%
  select(State, Variety) %>%
  left_join(tot_yields) %>%
  ggplot(aes(x = Year, y = Yield, color = Variety)) +
  geom_line() +
  facet_wrap(~State, ncol = 1)
```

## Final yields

```{r final_yields}
keys <- group_by(tot_yields, Group) %>%
  group_keys %>%
  pull

lapply(keys, function(gk) {
  vn <- sprintf("hops_%s_yields", tolower(gk))
  yd <- filter(tot_yields, Group == gk) %>% 
    select(-Group)
  fwrite(yd, file = here(SUBDIR, paste0(vn, ".csv")))
  assign(vn, value = yd, inherits = T)
})
```

[Back to top](#top)

## Summary

We now have multiple, short time-series to analyze at region and state levels. We also have more granular series by variety within Washington and Oregon, some of which are complete relative to state and region aggregates.

* Most HGA series are more complete than those obtained through NASS and are reasonably well validated for overlapping years in both sources.
* Most yield series are much shorter than our handful of complete Oregon and Washington varieties, limiting our options in what is already expected to be a challenging analysis. Idaho has _no_ completely reported yield series for any variety since 2000.
* Validation gives credibility to early sequences in the HGA series that could not be tested against NASS.

Examine the counts of yields by source and grouping.

```{r}
xtabs(~Group + Source, 
      data = mutate(tot_yields, Source = gsub("^HGA.*", "HGA", Source)))
```

Even after accounting for revisions to HGA yield tabulations that we corrected with NASS, the majority of validated yields are produced from HGA statistical reports.

### Challenges with yield data

For climatic and environmental analysis, For varietal analysis, hop yields can be more useful than acres harvested or tons produced. Unlike acreage, yield is a measure of areal density that, assuming efficient cultivation, removes demand or popularity of a variety from the analysis. But, hop yields are not always straightforward to interpret in modeling or other analysis.

1. _There is a practical upper limit to the yield that remains unknown from these data._ There is a limit to how many plants can be cultivated per acre from the space each plant needs to grow, and possibly the preferred spacing between plants to facilitate harvest. The number of cones produced per bine may vary naturally by hop variety.

2. _Efficiency of production is unknown._ Comparing yields across varieties or growing years assumes that their differences are due to natural variations, but independent of (unknowable to us) operational efficiency factors such as plant spacing, soil conditioning, or irrigation amounts. We have to assume for this analysis that each hop acre is cultivated as efficiently as any other regardless of variety or environment.

3. _The definition of yield requires clarification._ Hop varieties are often selected for the particular alpha-acid content for brewing as a proportion of the gross weight of the harvest. So alpha yield will differ from gross yield by chemical compounds sensitive to variety and likely cultivation practices. Another question is whether yields are what was harvested or what is commercially viable. To understand environmental impacts, we might design different analyses for different yield targets. This won't affect analysis of gross harvests, but should be kept in mind when reviewing yield analysis in the literature.

[Back to top](#top)

## Session info and notes

```{r session}
sessionInfo()
```

### Updates

* Apr 2026, maintenance on output paths, minor edits, and dropped remaining unevaluated portions of the script

* Sept 2025, major rewrite of scripting.

    * Added verification of NASS yields
    
    * Rewrote large portions of the scripting to automate error management and integration of HGA and NASS data. Eliminated manual replacement of missing HGA yields with error analysis and filtering with anti-join.

    * Labeling discrepancies with the Hallertauer varieties within HGA reporting and across the combined data sets are resolved.
    
    * Generalized the outlier analysis by region and variety, but eliminated the creation of dummy variables for each outlier.

### To Do

Nothing planned.

<!-- Write up comments on the unusually large discrepancies between HGA and NASS for some 2013 HGA reporting, and include an alternative yield series that replaces discrepant HGA values for NASS. -->

<!-- how is yield defined, by amount harvested (good or bad), sellable amount harvested, or something else? -->
